Themis-ml:

### 2017-09-06 paper/main.tex
complete draft of final paper submission

\documentclass{acm_proc_article-sp}
\usepackage{etoolbox}
\usepackage{graphicx}
\usepackage{hanging}
\usepackage{mdframed}
\usepackage{minted}
\usepackage{tabularx}

@@

@@

%\newpage
\section{Bias and Discrimination}
Colloquially, bias is simply a preference for or against something, e.g.,
preferring for vanilla over chocolate ice cream. While this definition is
intuitive, there are two types of bias that are less obvious and important to
highlight in this paper: algorithmic and machine learning bias.

Algorithmic bias is when a set of mathematical rules favors one set of
attributes over others in relation to some target variable, like ``approve'' or
``deny'' a loan. Machine learning bias is when a trained model systematically
generates predictions that favor one group over another in relation to some
set of attributes, e.g., education, and some target variable, e.g., ``default
on credit''. While these definitions of bias are amoral, discrimination is in
essence moral, occurring when an action is based on biases resulting in the
unfair treatment of people. We define fairness as the inverse of
discrimination, meaning that a ``fairness-aware'' model is one that produces
non-discriminatory predictions.

@@

\begin{table}
  \caption{A Simple Classification Pipeline}
  \renewcommand{\arraystretch}{1.75}
  \small\noindent\begin{tabularx}{\linewidth}{l X X}
    \textbf{API Interface} & \textbf{Function} & \textbf{E\(X_a\)mples} \\
    \hline
    \textbf{Transformer} &
      \emph{Preprocess} raw data for model training. &
      mean-unit variance scaling, min-max scaling \\
    \textbf{Estimator} &
      \emph{Train} models to perform a classification task. &
      logistic regression, random forest \\
    \textbf{Scorer} &
      \emph{Evaluate} performance of different models. &
      accuracy, f1-score, area under the curve \\
    \textbf{Predictor} &
      \emph{Predict} outcomes for new data. &
      single-classifier prediction, ensemble prediction
  \end{tabularx}
\end{table}

@@

The Scorer interface is concerned with measuring the degree to which data or
predictions are discriminatory. Themis-ml implements two methods for measuring
group-level discrimination and two methods for measuring individual-level
discrimination.

In the context of measuring group-level discrimination,
\cite{zliobaite2015survey} describes mean difference and normalized mean
difference. Mean difference measures the difference between \(p(a \cup y^{+})\)
and \(p(d \cup y^{+})\). Values range from -1 to 1, where -1 is the reverse-
discrimination case (all \(X_a\) have \(y^{-}\) labels and all \(X_d\) members
have \(y^{+}\) labels) and 1 is the fully discriminatory case (all \(X_a\) have
\(y^{+}\) labels and all \(X_d\) have \(y^{-}\) labels). Normalized mean
difference scales these values to 0 and 1 based on the maximum possible
discrimination in a dataset given the rate of positive labels
\cite{zliobaite2015survey}.

@@

\emph{\textbf{Work in progress}. Using the German Credit dataset, we evaluate
the efficacy of themis-ml in reducing discriminatory predictions. We do this by
comparing the performance of various fairness-aware techniques with a standard
``fairness- unaware'' classification system, as measured by the four
discrimination measures described above.}

@@

In this paper, we describe FMLIs in the classification context where we consider
only a single binary protected class variable. More work needs to be done to
generalize FMLIs to the multi-classification, regression, and multiple protected
classes contexts. Additionally, little is understood about the question of the
composability of fairness-aware methods. In other words, when different
techniques are used together in sequence, are the resulting discrimination
reductions additive or otherwise?

@@

% \begin{figure}[h]
% \centering
% \includegraphics[width=4cm]{IMG/d4gx-2015-logo.png}
% \caption{Conference logo for 2015 (PNG)}
% \label{d4gx-logo-2015}
% \end{figure}

% \begin{figure*}
%   \includegraphics[width=\textwidth]{IMG/d4gx-2015-logo.pdf}
%   \caption{Same logo, full width.}
% \end{figure*}

% \begin{figure}[h]
% \centering
% \includegraphics[width=7cm]{IMG/d4gx-2015-logo.pdf}
% \caption{Conference logo for 2015 (PDF)}
% \label{d4gx-logo-2015}
% \end{figure}

>>>>

\documentclass{acm_proc_article-sp}
\usepackage{etoolbox}
\usepackage{graphicx}
\usepackage{hanging}
\usepackage{mdframed}
\usepackage{minted}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage[labelfont=bf]{caption}
\captionsetup{labelfont = bf}

@@

\setlength{\parskip}{1em}

@@

%\newpage
\section{Bias and Discrimination}
Colloquially, bias is simply a preference for or against something, e.g.,
preferring for vanilla over chocolate ice cream. While this definition is
intuitive, here we explicitly define algorithmic bias as a form of bias that
occurs mathematical rules favors one set of attributes over others in relation
to some target variable, like ``approve'' or ``deny'' a loan.

Algorithmic bias in machine learning models can occur when a trained model
systematically generates predictions that favor one group over another in
relation to some set of attributes, e.g., education, and some target variable,
e.g., ``default on credit''. While the above definition of bias is amoral,
discrimination is in essence moral, occurring when an action is based on biases
resulting in the unfair treatment of people. We define fairness as the inverse
of discrimination, meaning that a ``fairness-aware'' model is one that produces
non-discriminatory predictions.

@@

\begin{table}
  \caption{A Simple Classification Pipeline}
  \renewcommand{\arraystretch}{1.75}
  \small\noindent\begin{tabularx}{\linewidth}{l X X}
    \textbf{API Interface} & \textbf{Function} & \textbf{Examples} \\
    \hline
    \textbf{Transformer} &
      \emph{Preprocess} raw data for model training. &
      mean-unit variance scaling, min-max scaling \\
    \textbf{Estimator} &
      \emph{Train} models to perform a classification task. &
      logistic regression, random forest \\
    \textbf{Scorer} &
      \emph{Evaluate} performance of different models. &
      accuracy, f1-score, area under the curve \\
    \textbf{Predictor} &
      \emph{Predict} outcomes for new data. &
      single-classifier prediction, ensemble prediction
  \end{tabularx}
\end{table}

@@

The Scorer interface is concerned with measuring the degree to which data or
predictions are potentially discriminatory (PD). Themis-ml implements two
methods for measuring group-level discrimination and two methods for measuring
individual-level discrimination.

In the context of measuring group-level discrimination,
\cite{zliobaite2015survey} describes mean difference and normalized mean
difference. Mean difference measures the difference between \(p(a \cup y^{+})\)
and \(p(d \cup y^{+})\). Values range from -1 to 1, where -1 is the reverse-
discrimination case (all \(X_a\) have \(y^{-}\) labels and all \(X_d\) members
have \(y^{+}\) labels) and 1 is the fully discriminatory case (all \(X_a\) have
\(y^{+}\) labels and all \(X_d\) have \(y^{-}\) labels). Normalized mean
difference, which also takes on values between -1 and 1, scales these values
based on the maximum possible discrimination in a dataset given the rate of
positive label \cite{zliobaite2015survey}.

@@

In this section we use the German Credit dataset \cite{bache2013uci}, mean
difference as the ``fairness'' measure, and area under the ROC curve (AUC) as
the ``utility'' measure. The following analysis is by no means meant to be a
comprehensive investigation of all of the possible workflows that themis-ml
enables. Rather, it is intended to demonstrate the potential of themis-ml as a
tool that facilites fairness-aware machine learning by enabling the user to:

\begin{enumerate}
  \item measure PD target label distributions in the training data.
  \item measure PD predictions in a machine learning algorithm's predictions.
  \item reduce PD predictions using fairness-aware techniques.
  \item diagnose the fairness-utility tradeoff in a particular data context.
\end{enumerate}

The German Credit dataset classifies 1000 anonymized individuals as having
``good'' and ``bad'' credit risks as part of a bank loan application, which we
encode as (\(1\)) and (\(0\)) respectively to define the \emph{credit risk}
target variable.

Each individual is associated with twenty attributes such as the purpose of the
loan, employment status, and other personal information. We begin the analysis
by extracting three protected class attributes --- \emph{female},
\emph{foreign worker}, and \emph{age below 25} --- and encode them as binary
variables such that the putatively disadvantaged group is encoded as \(1\) and
the advantaged group is encoded as \(0\) (the advantaged group would be
\emph{male}, \emph{citizen worker}, and \emph{age above 25} respectively).

Using the \textbf{Scorer} interface, we measure PD patterns with respect to
\emph{credit risk} and each of the protected classes defined above using the
\emph{mean difference} and \emph{normalized mean difference} metrics.

\textbf{Table 3} shows the PD distribution of ``good'' and ``bad'' credit risks
with respect to the protected attributes \emph{female}, \emph{foreign
worker}, and \emph{age below 25}. The mean difference (md) and normalized mean
difference (nmd) scores greater than zero suggests that the probability
of being classified as having ``good'' risk is higher in the advantaged group
than that of the disadvantaged group.

\begin{table}
  \caption{Potentially discriminatory target variable distribution.
    \emph{md} = mean difference, \emph{nmd} = normalized mean difference.}
  \renewcommand{\arraystretch}{1.75}
  \small\noindent\begin{tabularx}{\linewidth}{l|X|X|X|X}
    \textbf{protected class} & \textbf{md (\%)} & \textbf{md 95\% CI} &
      \textbf{nmd (\%)} & \textbf{nmd 95\% CI}\\
    \hline
    \textbf{female} & 7.48 & (1.35, 13.61) & 7.73 & (1.39, 14.06) \\
    \textbf{foreign worker} & 19.93 & (4.91, 34.94) & 63.96 & (15.76, 112.17)\\
    \textbf{age below 25} & 14.94 & (7.76, 22.13) & 17.29 & (8.97, 25.61)\\
  \end{tabularx}
\end{table}

\subsection{Experimental Procedure}

To assess the extent to which (i) a model trained on these data mirrors these PD
\emph{credit risk} distributions and (ii) fairness-aware techniques can reduce
these methods, we used \emph{mean difference} to measure model fairness and
\emph{AUC} to measure model utility. For this experiment we specify five
conditions:

\begin{itemize}
  \item Baseline (\emph{B}): Train a model on all available input variables in
        the German Credit dataset, including protected attributes.
  \item Remove Protected Attribute (\emph{RPA}): Train a model on input variables
        without protected attributes. This is the naive fairness-aware approach.
  \item Relabel Target Variable (\emph{RTV}): Train a model using the
        \emph{relabelling} fairness-aware method.
  \item Counterfactually Fair Model (\emph{CFM}): Train a model using the
        \emph{additive counterfactually fair} method.
  \item Reject-option Classification (\emph{ROC}): Train a model using the
        \emph{reject-option classification} method.
\end{itemize}

For each of these conditions, we train LogisticRegression, DecisionTree, and
RandomForest model types using 10-fold cross validation, generate train and test
predictions, and compute \emph{AUC} and \emph{mean difference} metrics for each
train-test pair. We then compute the mean these metrics for each condition and
model type.

\subsection{Measuring and Mitigating Potentially Discriminatory Predictions}

\begin{figure}[h]
\centering
\includegraphics[width=8.5cm]{IMG/fairness_aware_comparison.png}
\caption{
  \textbf{Comparison of Fairness-aware Methods} using LogisticRegression,
  DecisionTree, and RandomForest as base estimators as measured by AUC and
  mean difference evaluated on test set predictions.}
\label{figure-1-comparison-fa-methods}
\end{figure}

\textbf{Figure 1} suggests that in the case of LogisticRegression, the baseline
model \emph{B} does indeed mirror the PD patterns found in the true target
variable. Furthermore, each of the fairness-aware method appear to have
the desired effect of reducing \emph{mean difference}, but to variance degrees
depending on the method and protected attribute. In the \emph{female} context,
where there appear to be the least PD (mean difference of \(7.48\%\)), the
reductive effect of the fairness-aware methods does not appear to be as
large as in the \emph{foreign worker} and \emph{age below 25} contexts.

The lack of reduction in \emph{mean difference} between \emph{B} and \emph{RPA}
with respect to the \emph{foreign worker} protected attribute and
LogisticRegression model illustrates the observation made by
\cite{kusner2017counterfactual} that removing protected attributes from the
training data does not necessarly prevent the learning algorithm address from
mirroring PD patterns in the data.

However, as the sizeable reduction in \emph{mean difference} between \emph{B}
and \emph{RPA} with respect to the \emph{age below 25} protected attribute and
LogisticRegression model shows, removing protected attributes can make models
more fair while also retaining predictive power.

An interesting thing to note here is that the \emph{additive counterfactually
fair} method actually increases \emph{mean difference} for DecisionTrees and
RandomForests across all protected attribute contexts. A possible explanation
behind this observation is that certain assumptions made by the \emph{additive
counterfactually fair} method are not suitable for non-linear learning
algorithms, but this is an open question worth future inquiry.

\subsection{The Fairness-utility Tradeoff}

Just as the bias-variance tradeoff has become a useful diagnostic tool to guide
machine learning research and application, the fairness-utility tradeoff can
help machine learning practitioners and researchers understand how effective
particular fairness-aware methods might be in particular data contexts.

\begin{figure}[h]
\centering
\includegraphics[width=8.5cm]{IMG/fairness_utility_tradeoff.png}
\caption{
  \textbf{Correlation between AUC and Mean Difference} for each experimental
  condition, across all model types (LogisticRegression, DecisionTree,
  RandomForest). 95\% confidence intervals are provided for the pearson \(R\)
  correlation metric.
  }
\label{figure-2-fairness-utility-tradeoff}
\end{figure}

\textbf{Figure 2} demonstrates an initial attempt at visualizing the the
fairness-utility tradeoff, in this case as measured by \emph{mean difference}
and \emph{AUC} respectively. These results suggest that the relationship between
fairness and utility is noisy, however there does seem to be a positive
correlation between \emph{mean difference} and \emph{AUC}, or a negative
correlation between fairness and utility (since lower scores are better
for \emph{mean difference} and higher scores are better for \emph{AUC}).

Depending on one's use cases, analyses like this might be prove to be a useful
guide for figuring out what kinds of methods are robust in the sense that
one can reduce PD predictions with little to no adverse impact on predictive
performance.

@@

In this paper, we describe and evaluate an FMLI in the classification context
where we consider only a single binary protected class variable and a binary
target variable. More work needs to be done to generalize FMLIs to the multi-
classification, regression, and multiple protected classes contexts.

Furthermore, many basic questions about model tuning, model evaluation, and
model selection in relation to the fairness-utility tradeoff remain open. For
instance, what might be some reasonable ways to aggregate utility and
performance metrics in order to find the optimal set of hyperparameters?
Additionally, little is understood about the question of the composability of
fairness-aware methods. In other words, when different techniques are used
together in sequence, are the resulting discrimination reductions additive or
otherwise?

@@
